Finally Tagless, Partially Evaluated

### 2008-10-03 tagless/tagless-final.tex
Repack related work

Figure~\ref{fig:pe} shows the partial evaluator~|P|.
Its type |repr| expresses the definition for |static| given
at the start of this section, with |'sv| taking the crucial place of |('c,'dv) static|.
The function |abstr|
extracts a future-stage code value from the result of
partial evaluation.  Conversely, the function |pdyn| injects a
code value into the |repr| type. As
in~\S\ref{S:PE-problem}, we build dynamic terms alongside
any static ones to express how the lift function is indexed
by the dynamic type.  What is new is that we analogously
build a static type alongside the dynamic type to express
how the static type is indexed by the dynamic type.

@@

Predating us, \citet{Thiemann-combinators} and \citet{sumii-hybrid} deforested
the object term representation and expressed a partial evaluator as a collection
of term combinators in a typed metalanguage.  Like us, \citeauthor{sumii-hybrid}
follow \citet{Sperber-SelfApplicable} and \citet{asai-binding-time} in building
static and dynamic terms in tandem, to combine offline and online partial
evaluation.
\citearound{'s earlier self-reducer for the untyped
$\lambda$-calculus}\Citet{JFP-Mogensen} also
builds static and dynamic terms in tandem.  However, it
builds a static term for every object term, even a bound variable,
so it moves some work from |app| to~|pdyn| (in terms of Figure~\ref{fig:pe})
and remains untyped.  In contrast, we follow Sperber, Asai,
and \citeauthor{sumii-hybrid} in leaving the static term
optional, so as to perform lifting without
juggling explicit type indices in the encoding of an object term.

Our contribution to the literature on partial evaluation is to use mere
Hindley-Milner types in the metalanguage to assure statically and patently that
partially evaluating a well-typed object program not only never gets stuck but
also, if it terminates, produces a well-typed output program that never gets
stuck.  Moreover, these types form an instance of a general |Symantics|
signature that encompasses other interpreters such as evaluation and
compilation.  This early and manifest assurance of type safety contrasts, for
example, with \citearound{'s compiler generator (cogen) for
ML}\citet{Birkedal-PE-ML}, which transforms a program into its tagless
generating extension.  Because that cogen uses a universal type, the fact that
it never generates an ill-typed generating extension from a well-typed input
program is only manifest when each generating extension is type-checked, and the
fact that the generating extension never generates an ill-typed residual program
from well-typed static input is only manifest when each residual program is
type-checked.  Similarly, the fact that the partial evaluator of
\citet{fiore:nbe-ppdp2002}, nor that of
\citet{balat:tdpe-popl2004}, both of which use delimited control
operators, never turn well-typed code into ill-typed code is not assured by the
metalanguage, whether or not as part of a typed family of interpreter modules.
\begin{comment}
% Removed the following comment in view of Section 4.5 of Danvy's
% original TDPE paper (POPL 1996)
It is
type-directed, so the user must represent, as a term, the type of every
term to be partially evaluated.  We shift this work to the type checker
of the metalanguage.
By avoiding term-level type representations, our
approach makes it easier to perform algebraic simplifications (as
in~\S\ref{S:PE-solution}).
\end{comment}

@@

Even a simply typed $\lambda$-calculus obviously supports self\hyp
interpretation, provided we use universal types \citep{taha-tag}.  The
ensuing tagging overhead motivated \citet{Makholm-TagElim,taha-tag} 
to propose tag
elimination, which however does not statically guarantee that all tags
will be removed \citep{WalidICFP02}.

\Citet{WalidICFP02}, \citet{taha-tag}, \citet{xi-guarded}, and
\citet{peyton-jones-simple} seem to argue as follows that a self\hyp
interpreter of a typed language cannot be tagless or Jones\hyp optimal:
\begin{shortlist}
\item One needs to encode a typed language in a typed language based on
a sum type (at some level of the hierarchy)\\
\item A \emph{direct} interpreter 
for such an encoding of a typed language
in a typed language requires either
advanced types or tagging overhead\\
\item Thus, an indirect interpreter is necessary, which needs a universal
  type and hence tagging\ifshort\else\\
\item Thus, any self-interpreter must have tags and cannot be 
  Jones-optimal\fi.
\end{shortlist}
While the logic is sound, we (following \citet{yang-encoding}) showed that the
first step's premise is not valid.

@@

That it is possible to regard our various evaluators as forming a 
family is also an old idea \citep{Holst-AMIX}. \Citeauthor{Holst-AMIX}
introduces \emph{language triplets} as an interesting mechanism to 
relate (untyped) languages in much the same way we do.  Seeds of 
several of the techniques we use here can also be found in this paper.

Our partial evaluator establishes a bijection |static| between static
and dynamic types (the valid values of |'sv| and |'dv|), and between
static and dynamic terms.  It is customary to implement such a bijection
using an injection\hyp projection pair, as done for interpreters
\ifshort \citep{Ramsey-ML-module-mania,Benton-embedded-interpreters}\else
by \citet{Ramsey-ML-module-mania} and \citet{Benton-embedded-interpreters}\fi,
partial evaluation \ifshort \citep{Danvy-TDPE}\else by \citet{Danvy-TDPE}\fi,
and type-level functions \ifshort \citep{oliveira-typecase}\else by
\citet{oliveira-typecase}\fi.  As explained in~\S\ref{S:PE-solution}, we
avoid injection and projection at the type level by adding an argument
to |repr|.
Our solution could have been even more straightforward if MetaOCaml
provided total type-level functions such as |static| in
\S\ref{S:PE-solution}---simple type-level computations 
ought to become mainstream.

Our separation between the |Symantics| interface and its many
implementations codifies the common practice of implementing an embedded
DSL by specifying an abstract syntax of object\hyp language pervasives,
such as addition and application, then providing multiple
interpretations of them.  \Citet{JonesNielson:94:AbstractInterpretation}
prefigured this separation when they decomposed a denotational definition of an
untyped object language into a core semantics (which we call abstract syntax)
and multiple interpretations.  When implementing an embedded DSL in this
way, it is also common practice to use phantom types to rule out
ill-typed object terms, as done in Lava
\citep{Lava} and by \citet{Rhiger-thesis}. However, these two approaches
are not tagless because they still use universal types, such as Lava's
\texttt{Bit} and \texttt{NumSig}, and Rhiger's \texttt{Raw} (his Figure~2.2)
and \texttt{Term} (his Chap.~3), which incur the attendant overhead of
pattern matching.  The universal type also
greatly complicates the soundness and completeness proofs of embedding
\citep{Rhiger-thesis}, whereas our proofs are trivial.
Rhiger's approach does not support typed CPS transformation (his~\S3.3.4).
\begin{comment}
Rhiger's But Fig 2.2, p33: universal type Raw.  He uses phantom type
upon the Exp datatype. But that is cheating: phantom type means
essentially we can easily do coerce. We use real types.  That's why he
had to do tedious proofs in Sec 2 of soundness and completeness of
embedding. Whereas our proofs are obvious.  His sec 3 is based on data
representation of terms. They have type tags.  We do nothing of that
kind: See Sec 3.1.2. See numerous "data Term" in Sec3, which is the U
type.  In Sec 3.3.4 (p76) Rhiger specifically says that his encoding
cannot do typed CPS transformation -- whereas our does. BTW, Rhiger
thesis contains the definitions of the interpreter and the compiler,
in the beginning. Use this in response to Rev1
\end{comment}

In many ways, our work builds on that of \citep{Thiemann-combinators}.
We use the same basic techniques, but our partial evaluator is online
as this is the only way to make partial evaluation ``fit'' in our
family.  In other words, by moving to an online partial evaluator and
using staging instead of a $2$-level language, we are able to show that
\citeauthor{Thiemann-combinators}'s techniques are highly parametric
and typeable using mainstream features.  Using many different type
classes, including constructor classes, \citep{Thiemann-GenClass}
presents a Gofer implementation of set of \emph{binding-time polymorphic}
combinators.  By gathering all their classes into one and dropping the
requirement for a polymorphic |lift|, they could have invented
|Symantics| themselves.

@@

Our encoding of the type function |static| in \S\ref{S:PE-solution}
emulates type-indexed types and is related to intensional type analysis
\citep{Morrisett-intensional,Generic-Haskell}. However, our object
language and running examples in HOAS include |fix|,
which intensional type analysis cannot handle
\citep{xi-guarded}.

>>>>

Figure~\ref{fig:pe} shows the partial evaluator~|P|.
Its type |repr| expresses the definition for |static| given
at the start of this section, with |'sv| taking the crucial place of |('c,'dv) static|.
The function |abstr|
extracts a future-stage code value from the result of
partial evaluation.  Conversely, the function |pdyn| injects a
code value into the |repr| type.
Thus, |abstr| and |pdyn| are like the \emph{reify} and
\emph{reflect} functions defined in normalization by evaluation
\citep{Danvy-TDPE}, but as
in~\S\ref{S:PE-problem}, we build dynamic terms alongside
any static ones to express how the lift function is indexed
by the dynamic type.  Analogously, we now
build a static type alongside the dynamic type to express
how the static type is indexed by the dynamic type.
Thus we establish a bijection |static| between static and dynamic types,
without defining at the type level the injection\hyp projection pairs
customarily used to establish such bijections for interpreters
\citep{Ramsey-ML-module-mania,Benton-embedded-interpreters},
partial evaluation \citep{Danvy-TDPE},
and type-level functions \citep{oliveira-typecase}.
This emulation of type-indexed types is related to intensional type analysis
\citep{Morrisett-intensional,Generic-Haskell}, but
intensional type analysis cannot handle our |fix|
\citep{xi-guarded}.

@@

Our partial evaluator owes much to
\citet{Thiemann-combinators} and \citet{sumii-hybrid}, who deforested
the object term representation and expressed a partial evaluator as a collection
of term combinators in a typed metalanguage.  Like us, \citeauthor{sumii-hybrid}
follow \citet{Sperber-SelfApplicable} and \citet{asai-binding-time} in building
static and dynamic terms in tandem, to combine offline and online partial
evaluation.
\citearound{'s earlier self-reducer for the untyped
$\lambda$-calculus}\Citet{JFP-Mogensen} also
builds static and dynamic terms in tandem.  However, it
builds a static term for every object term, even a bound variable,
so it moves some work from |app| to~|pdyn| (in terms of Figure~\ref{fig:pe})
and remains untyped.  In contrast, we follow Sperber, Asai,
and \citeauthor{sumii-hybrid} in leaving the static term
optional, so as to perform lifting without
juggling explicit type indices in the encoding of an object term.

Our contribution to the literature on partial evaluation is to use mere
Hindley-Milner types in the metalanguage to assure statically and patently that
partially evaluating a well-typed object program not only never gets stuck but
also, if it terminates, produces a well-typed output program that never gets
stuck.  Moreover, thanks to the online binding-time analysis performed by
our partial evaluator (in contrast to \citeauthor{Thiemann-combinators}'s),
these types form an instance of a general |Symantics|
signature that encompasses other interpreters such as evaluation and
compilation.  This early and manifest assurance of type safety contrasts, for
example, with \citearound{'s compiler generator (cogen) for
ML}\citet{Birkedal-PE-ML}, which transforms a program into its tagless
generating extension.  Because that cogen uses a universal type, the fact that
it never generates an ill-typed generating extension from a well-typed input
program is only manifest when each generating extension is type-checked, and the
fact that the generating extension never generates an ill-typed residual program
from well-typed static input is only manifest when each residual program is
type-checked.  Similarly, the fact that the partial evaluator of
\citet{fiore:nbe-ppdp2002} and that of
\citet{balat:tdpe-popl2004}, both of which use delimited control
operators, never turn well-typed code into ill-typed code is not assured by the
metalanguage, whether or not as part of a typed family of interpreter modules.
\begin{comment}
% Removed the following comment in view of Section 4.5 of Danvy's
% original TDPE paper (POPL 1996)
It is
type-directed, so the user must represent, as a term, the type of every
term to be partially evaluated.  We shift this work to the type checker
of the metalanguage.
By avoiding term-level type representations, our
approach makes it easier to perform algebraic simplifications (as
in~\S\ref{S:PE-solution}).
\end{comment}

@@

The simply typed $\lambda$-calculus can be interpreted even just in
itself, provided we use universal types \citep{taha-tag}.  The
ensuing tagging overhead motivated \citet{Makholm-TagElim,taha-tag} 
to propose \emph{tag
elimination}, which however does not statically guarantee that all tags
will be removed \citep{WalidICFP02}.

\Citet{WalidICFP02}, \citet{taha-tag}, \citet{xi-guarded}, and
\citet{peyton-jones-simple} seem to argue that a typed
interpreter of a typed language cannot be tagless without
advanced types, based on the premise that the only way to
encode a typed language in a typed language is to use
a sum type (at some level of the hierarchy).
While the logic is sound, we (following \citet{yang-encoding}) showed that the
premise is not valid.

@@

Our separation between the |Symantics| interface and its many
implementations codifies the common practice of implementing an embedded
DSL by specifying an abstract syntax of object\hyp language pervasives,
such as addition and application, then providing multiple
interpretations of them.
The techniques we use to form such a family of interpreters
find their seeds in \citearound{'s \emph{language
triplets}}\citet{Holst-AMIX}, though in an untyped setting.
\Citet{JonesNielson:94:AbstractInterpretation} also
prefigured this separation when they decomposed a denotational definition of an
untyped object language into a core semantics (which we call abstract syntax)
and multiple interpretations.  When implementing an embedded DSL in this
way, it is also common practice to use phantom types to rule out
ill-typed object terms, as done in Lava
\citep{Lava} and by \citet{Rhiger-thesis}. However, these two approaches
are not tagless because they still use universal types, such as Lava's
\texttt{Bit} and \texttt{NumSig}, and Rhiger's \texttt{Raw} (his Figure~2.2)
and \texttt{Term} (his Chap.~3), which incur the attendant overhead of
pattern matching.  The universal type also
greatly complicates the soundness and completeness proofs of embedding
\citep{Rhiger-thesis}, whereas our proofs are trivial.
Rhiger's approach does not support typed CPS transformation (his~\S3.3.4).
\begin{comment}
Rhiger's But Fig 2.2, p33: universal type Raw.  He uses phantom type
upon the Exp datatype. But that is cheating: phantom type means
essentially we can easily do coerce. We use real types.  That's why he
had to do tedious proofs in Sec 2 of soundness and completeness of
embedding. Whereas our proofs are obvious.  His sec 3 is based on data
representation of terms. They have type tags.  We do nothing of that
kind: See Sec 3.1.2. See numerous "data Term" in Sec3, which is the U
type.  In Sec 3.3.4 (p76) Rhiger specifically says that his encoding
cannot do typed CPS transformation -- whereas our does. BTW, Rhiger
thesis contains the definitions of the interpreter and the compiler,
in the beginning. Use this in response to Rev1
\end{comment}

\Citet{Thiemann-GenClass}
implemented a set of \emph{binding-time polymorphic}
combinators in Gofer, using many constructor classes.
By merging all their classes into one and dropping
polymorphic lift, they could have invented
|Symantics|.

@@
